---
title: "Human FSI intrinsic ephys parameters analysis"
author: "Florian Wildner"
date: "14.04.2026"
output:
  html_document: default

# Version notes
version: "16.12 – inclusion of anatomy data"
last_changes: "Reordered protocols to align with manuscript structure"
---

\tiny
```{r echo=FALSE, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 5,
  fig.height = 4
)

# -----------------------------
# Libraries
# -----------------------------
library(umap)
library(ggcorrplot)
library(ggplot2)
library(ggpubr)
library(sigclust)
library(dendextend)
library(rstatix)
library(viridis)
library(gridExtra)
library(scico)
library(gghalves)
library(tidyverse)
library(dplyr)

# -----------------------------
# Output directory + save helper
# -----------------------------

# create date string (YYMMDD)
date_str <- format(Sys.Date(), "%y%m%d")

# relative output path (inside project folder)
out_dir <- file.path("output", date_str)

# create directory if it doesn't exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

options(fig_out_dir = out_dir)

save_fig <- function(plot, name, w = 5, h = 4, device = "eps") {

  out_dir <- getOption("fig_out_dir")

  ggsave(
    filename = file.path(out_dir, paste0(name, ".", device)),
    plot = plot,
    width = w,
    height = h,
    units = "in",
    device = device
  )
}
```
\large
# **General analysis approach**
\small
## 0.1 Load and prepare data
Raw electrophysiological data are loaded from CSV, filtered for cell identity, and split into metadata and numeric features. Quality control excludes recordings with depolarized resting membrane potentials.
\tiny
```{r}
# -----------------------------
# Load data
# -----------------------------
col <- c(
  "c","c","n","c","c",
  "c","c","c","c","c","c",
  "n","c","c",
  "n","n","n","n","n","n","n","n","n","n",
  "n","n","n","n","n","n","n","n","n","n",
  "n","n","n","n","n","n","n","n"
)

df_original <- read_csv(
  "/Users/flo/Desktop/research/Human/Human_data/intrinsic_properties/Human_IN_Properties_260124.csv",
  col_types = col
)

# df_original <- read_csv(
#   file.path("data", "Human_IN_Properties.csv"),
#   col_types = col
# )

# -----------------------------
# Filter datasets
# -----------------------------
df_sel <- df_original %>%
  filter(ident %in% c("WFA+", "DCF+"))

df_all <- df_original %>%
  filter(ident %in% c("WFA+", "DCF+", "none"))

df <- df_sel
# df <- df_all

# -----------------------------
# QC threshold
# -----------------------------
rmp_cutoff <- -50

# -----------------------------
# Split metadata vs numeric data
# -----------------------------
df_meta <- select(df, cell, pat, age, sex, patho, area, clinic,
                  transport, put_type, ident, burst)

df_anat_raw <- select(df, cell, anat_morph, anat_dist, anat_pv, anat_sst)

df_dbl <- select(df,
  cell, RMP, CC_Rm, rheobase, CC_tau, sag_ratio,
  freq_max, f_i_slope, onset_lat,
  isi_first, isi_avg, isi_cv, isi_adapt,
  amp_cv, amp_adapt,
  ap_amp, ap_thresh, ap_width,
  ap_rise, ap_decay,
  ap_upstroke, ap_downstroke, ap_UDR,
  fAHP_amp, fAHP_time
)

# -----------------------------
# Clean dataset
# -----------------------------
df_dbl <- df_dbl %>%
  drop_na() %>%
  filter(RMP <= rmp_cutoff)

cells <- select(df_dbl, cell)
df_dbl <- select(df_dbl, -cell)

```
\small
## 0.2 Setup visualization
1. Define consistent color palette (Okabe-Ito) for identification methods and clusters
2. Create parameter dictionary for publication-ready axis labels
\tiny
```{r}
# -----------------------------
# Color palette (Okabe–Ito)
# -----------------------------
my_colors <- c(
  "WFA+" = "#CC79A7",
  "DCF+" = "#94D4C6",
  "1" = "#56B4E9",
  "2" = "#E69F00",
  "3" = "#D55E00",
  "none" = "#999999",
  "classical FSI" = "#56B4E9",
  "non-FSI" = "#E69F00",
  "borderline-FSI" = "#D55E00"
)

# -----------------------------
# Publication label dictionary
# -----------------------------
param_dict <- c(
  amp_adapt = "Amplitude adaptation index",
  amp_cv = "Amplitude CV",
  ap_amp = "AP amplitude (mV)",
  ap_decay = "AP decay time (ms)",
  ap_downstroke = "AP downstroke (mV/ms)",
  ap_rise = "AP rise time (mV/ms)",
  ap_thresh = "AP threshold (mV)",
  ap_UDR = "AP upstroke/downstroke ratio",
  ap_upstroke = "AP upstroke (ms)",
  ap_width = "AP width (ms)",
  CC_Rm = "Input resistance (MΩ)",
  CC_tau = "Membrane tau (ms)",
  f_i_slope = "F/I slope (Hz/mV)",
  fAHP_amp = "fAHP amplitude (mV)",
  fAHP_time = "fAHP time (ms)",
  freq_max = "Max. firing rate (Hz)",
  isi_adapt = "ISI adaptation index",
  isi_avg = "ISI mean (ms)",
  isi_cv = "ISI CV",
  isi_first = "First ISI (ms)",
  onset_lat = "Onset latency (ms)",
  rheobase = "Rheobase (pA)",
  RMP = "Resting membrane potential (mV)",
  sag_ratio = "Sag ratio"
)

# -----------------------------
# Publication label dictionary
# without special characters
# -----------------------------

# Parameter dictionary without special characters
param_dict_safe <- c(
  amp_adapt    = "Amplitude adaptation index",
  amp_cv       = "Amplitude CV",
  ap_amp       = "AP amplitude (mV)",
  ap_decay     = "AP decay time (ms)",
  ap_downstroke= "AP downstroke (mV/ms)",
  ap_rise      = "AP rise time (mV/ms)",
  ap_thresh    = "AP threshold (mV)",
  ap_UDR       = "AP upstroke/downstroke ratio",
  ap_upstroke  = "AP upstroke (ms)",
  ap_width     = "AP width (ms)",
  CC_Rm        = "Input resistance (MOhm)",
  CC_tau       = "Membrane tau (ms)",
  f_i_slope    = "F/I slope (Hz/mV)",
  fAHP_amp     = "fAHP amplitude (mV)",
  fAHP_time    = "fAHP time (ms)",
  freq_max     = "Max. spiking frequency (Hz)",
  isi_adapt    = "ISI adaptation index",
  isi_avg      = "ISI average (ms)",
  isi_cv       = "ISI CV (ms)",
  isi_first    = "ISI first (ms)",
  onset_lat    = "Onset latency (ms)",
  rheobase     = "Rheobase (pA)",
  RMP          = "RMP (mV)",
  sag_ratio    = "Sag ratio (mV)"
)
```
\small
## 0.3 Quality control: Compare input resistance (VC_Rs) and RMP between methods
1. Boxplots + normality tests (Shapiro-Wilk) + appropriate statistical test (t-test/Wilcoxon)
2. Ensures comparable recording quality between DCF+ and WFA+ identified cells
\tiny
```{r}
# Subset the data for the two groups (DCF+ and WFA+)
df_subset <- df %>%
  filter(ident %in% c("DCF+", "WFA+")) %>%
  filter(RMP <= rmp_cutoff)

# VC_Rs comparison
ggplot(df_subset, aes(ident, VC_Rs, fill = ident)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.1), size = 1) +
  scale_fill_manual(values = my_colors) +
  theme_minimal() +
  labs(x = "Group", y = "VC_Rs")

# Normality tests
shapiro_dcf <- shapiro.test(df_subset$VC_Rs[df_subset$ident == "DCF+"])
shapiro_wfa <- shapiro.test(df_subset$VC_Rs[df_subset$ident == "WFA+"])

if (shapiro_dcf$p.value > 0.05 & shapiro_wfa$p.value > 0.05) {
  t.test(VC_Rs ~ ident, data = df_subset)
} else {
  wilcox.test(VC_Rs ~ ident, data = df_subset)
}

# Same for RMP
ggplot(df_subset, aes(ident, RMP, fill = ident)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.1), size = 1) +
  scale_fill_manual(values = my_colors) +
  theme_minimal()

# Shapiro-Wilk normality test for each group
shapiro_dcf <- shapiro.test(df_subset$RMP[df_subset$ident == "DCF+"])
shapiro_wfa <- shapiro.test(df_subset$RMP[df_subset$ident == "WFA+"])

# Check if data is normally distributed
if (shapiro_dcf$p.value > 0.05 & shapiro_wfa$p.value > 0.05) {
  # Perform a t-test if data is normally distributed
  t_test_result <- t.test(RMP ~ ident, data = df_subset)
  print("T-test result for RMP (DCF+ vs. WFA+):")
  print(t_test_result)
} else {
  # Perform a non-parametric Wilcoxon test if data is not normally distributed
  wilcox_test_result <- wilcox.test(RMP ~ ident, data = df_subset)
  print("Wilcoxon test result (non-parametric test) for RMP (DCF+ vs. WFA+)::")
  print(wilcox_test_result)
}
```
\large
# 1.: Comparing identification methods (DCF vs WFA) (Fig 1)
MANOVA → univariate tests → pairwise comparisons
1. MANOVA between WFA and DCF\
2. univariate tests
3. pairwise comparisons
\small
## 1.1: Pre-process for analysis: create data frame df_int 
```{r}
# Initialize dataframe
df_int <- select(df,cell,ident,put_type,
                RMP,CC_Rm,CC_tau,sag_ratio,rheobase,
                freq_max, f_i_slope, onset_lat,
                isi_first, isi_avg, isi_cv, isi_adapt,
                amp_cv, amp_adapt,
                ap_amp, ap_thresh, ap_width,
                ap_rise, ap_decay, 
                ap_upstroke, ap_downstroke, ap_UDR, 
                fAHP_amp, fAHP_time) 
df_int<-drop_na(df_int)%>%
  filter(RMP <= rmp_cutoff)  # Exclude RMP > cutoff
# Ensure ident is a factor
df_int$ident <- factor(df_int$ident, levels = c("DCF+", "WFA+"))
```
\small
## 1.2: Comparison of electrophysiological parameters between DCF+ and WFA+
1. A multivariate analysis of variance (MANOVA) was used to test for overall group differences across all 24 electrophysiological parameters between DCF+ and WFA+ neurons.
2. If the MANOVA was significant, univariate follow-up tests were performed for each parameter:
   - Normality was assessed per parameter within each group (Shapiro–Wilk test)
   - Depending on distribution, either an independent t-test (parametric) or Wilcoxon rank-sum test (non-parametric) was applied
   - P-values were corrected for multiple comparisons using the Benjamini–Hochberg false discovery rate (FDR)
   - Effect sizes were calculated using Cohen’s d
3. Visualization:
   - Significant parameters (FDR-corrected) were visualized
   - A predefined subset of key physiological parameters was additionally plotted (**Figure 1C**)
4. A summary table including all parameters, statistical results, and effect sizes was generated for the supplement (**Supplementary Table 1**)
```{r}
# --- Prepare data for MANOVA ---
df_manova <- df_int %>%
  select(ident,
         RMP, CC_Rm, CC_tau, sag_ratio, rheobase,
         freq_max, f_i_slope, onset_lat,
         isi_first, isi_avg, isi_cv, isi_adapt,
         amp_cv, amp_adapt,
         ap_amp, ap_thresh, ap_width,
         ap_rise, ap_decay,
         ap_upstroke, ap_downstroke, ap_UDR,
         fAHP_amp, fAHP_time) %>%
  drop_na()

# Get parameter names
param_names <- setdiff(names(df_manova), "ident")

cat("Sample sizes:\n")
print(table(df_manova$ident))

# --- STEP 1: PRIMARY TEST - MANOVA ---
cat("\n=== MANOVA: Overall Multivariate Test ===\n")
cat("Null hypothesis: WFA+ and DCF+ have identical multivariate means across all 24 parameters\n\n")

manova_model <- manova(as.matrix(df_manova[, param_names]) ~ ident, 
                       data = df_manova)

# Use Pillai's trace
manova_summary <- summary(manova_model, test = "Pillai")
print(manova_summary)

# Extract p-value
manova_p <- manova_summary$stats[1, "Pr(>F)"]

# --- STEP 2: FOLLOW-UP UNIVARIATE TESTS (only if MANOVA significant) ---
if (manova_p < 0.05) {
  
  cat("\n=== MANOVA is significant (p < 0.05) ===")
  cat("\nProceeding to univariate tests to identify which parameters differ...\n\n")
  
  # Reshape for testing
  df_long <- df_manova %>%
    pivot_longer(cols = all_of(param_names), 
                 names_to = "Parameter", 
                 values_to = "Value")
  
  # Normality check per parameter per group
  normality <- df_long %>%
    group_by(Parameter, ident) %>%
    summarise(
      n = n(),
      p.normality = if(n() >= 3) shapiro.test(Value)$p.value else NA_real_,
      .groups = "drop"
    ) %>%
    mutate(normal = p.normality > 0.05 | is.na(p.normality))
  
  # Identify which parameters need non-parametric tests
  non_normal_params <- normality %>%
    group_by(Parameter) %>%
    summarise(any_non_normal = any(!normal, na.rm = TRUE)) %>%
    filter(any_non_normal) %>%
    pull(Parameter)
  
  cat("\nParameters requiring non-parametric test:", 
      length(non_normal_params), "of", length(param_names), "\n")
  
  # Parametric tests (t-test)
  stat_test_t <- df_long %>%
    filter(!Parameter %in% non_normal_params) %>%
    group_by(Parameter) %>%
    t_test(Value ~ ident) %>%
    mutate(method = "t-test")
  
  # Non-parametric tests (Wilcoxon)
  stat_test_w <- df_long %>%
    filter(Parameter %in% non_normal_params) %>%
    group_by(Parameter) %>%
    wilcox_test(Value ~ ident) %>%
    mutate(method = "Wilcoxon")
  
  # Combine results
  stat_test_combined <- bind_rows(stat_test_t, stat_test_w)
  
  # Apply FDR correction (Benjamini-Hochberg)
  stat_test_fdr <- stat_test_combined %>%
    ungroup() %>%
    mutate(p.adj.fdr = p.adjust(p, method = "BH")) %>%
    mutate(p.adj.signif.fdr = case_when(
      p.adj.fdr <= 0.001 ~ "***",
      p.adj.fdr <= 0.01  ~ "**",
      p.adj.fdr <= 0.05  ~ "*",
      TRUE ~ "ns"
    )) %>%
    arrange(p.adj.fdr)
  
  # Also calculate Bonferroni for comparison
  stat_test_fdr <- stat_test_fdr %>%
    mutate(p.adj.bonf = p.adjust(p, method = "bonferroni")) %>%
    mutate(p.adj.signif.bonf = case_when(
      p.adj.bonf <= 0.001 ~ "***",
      p.adj.bonf <= 0.01  ~ "**",
      p.adj.bonf <= 0.05  ~ "*",
      TRUE ~ "ns"
    ))
  
  # --- STEP 3: EFFECT SIZES ---
  effect_sizes <- df_long %>%
    group_by(Parameter) %>%
    cohens_d(Value ~ ident) %>%  # Removed pooled_sd argument
    select(Parameter, effsize, magnitude)
  
  # Combine all results
  final_results <- stat_test_fdr %>%
    left_join(effect_sizes, by = "Parameter") %>%
    select(Parameter, method, statistic, p, p.adj.fdr, p.adj.signif.fdr, 
           p.adj.bonf, p.adj.signif.bonf, effsize, magnitude)
  
  cat("\n=== Results Summary ===\n")
  print(final_results, n = 24)
  
  
  # --- STEP 4: VISUALIZATION ---
  # Create a function for plotting in the same style
  plot_ephys_params <- function(df_long, final_results, parameters_to_plot) {
    # Filter data to selected parameters
    parameters_to_plot <- factor(parameters_to_plot, levels = parameters_to_plot, ordered = TRUE)
    df_plot <- df_long %>%
      filter(Parameter %in% parameters_to_plot) %>%
      mutate(Parameter = factor(Parameter, levels = parameters_to_plot, ordered = TRUE))
    
    # Also ensure parameters_to_plot is a character vector for consistency
    parameters_to_plot <- as.character(parameters_to_plot)
    
    # Calculate y positions for significance bars (only for parameters present in final_results)
    y_positions <- df_plot %>%
      group_by(Parameter) %>%
      summarize(y.position = max(Value, na.rm = TRUE) + 0.15 * abs(max(Value, na.rm = TRUE)), .groups = "drop")
    
    # Prepare stat annotation subset to selected parameters
    stat_anno <- final_results %>%
      filter(Parameter %in% parameters_to_plot) %>%
      left_join(y_positions, by = "Parameter") %>%
      mutate(
        group1 = "DCF+",
        group2 = "WFA+",
        label = p.adj.signif.fdr
      )
    
    # Create plot 
    p <- ggplot(df_plot, aes(x = ident, y = Value, color = ident)) +
      coord_cartesian(clip = "off") +
      geom_jitter(width = 0.1, size = 0.6, alpha = 1, na.rm = TRUE) +
      geom_violin(width = 0.7, trim = TRUE, alpha = 0.25, linewidth = 0.3, na.rm = TRUE) +
      geom_boxplot(width = 0.15, outlier.shape = NA, color = "black", alpha = 0.3, size = 0.3, na.rm = TRUE) +
      stat_pvalue_manual(stat_anno, label = "label", tip.length = 0.01, size = 3, hide.ns = TRUE) +
      scale_color_manual(values = my_colors) +
      facet_wrap(~factor(Parameter, levels = parameters_to_plot), 
                 scales = "free_y", ncol = 3, 
                 labeller = as_labeller(param_dict), drop = FALSE, as.table = TRUE) +
      labs(title = NULL, x = NULL, y = NULL) +
      theme_classic(base_size = 14) +
      theme(
        panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 6),
        axis.text.y = element_text(size = 6, color = "black"), 
        axis.text = element_text(color = "black"),
        axis.line = element_line(color = "black"),
        strip.background = element_blank(),
        strip.text = element_text(size = 8, face = "plain"),
        legend.position = "none",
        plot.margin = margin(t = 15, r = 5, b = 5, l = 5)
      )
    
    return(p)
  }
  
  # 1. Plot only significant parameters from MANOVA
  sig_params <- final_results %>% filter(p.adj.fdr < 0.05) %>% pull(Parameter)
  p_sig <- plot_ephys_params(df_long, final_results, sig_params)
  print(p_sig)
  
  # 2. Plot selected, typical parameters regardless of significance
  custom_params <- c(
      "ap_width", "ap_upstroke", "fAHP_amp", "ap_thresh", 
      "freq_max", "rheobase", "amp_adapt", "isi_adapt", 
      "sag_ratio", "CC_Rm", "CC_tau", "RMP"
      )
  p_custom <- plot_ephys_params(df_long, final_results, custom_params)
  print(p_custom)
  save_fig(p_custom, "Fig_1C_Ephys_custom_parameters", w=280/96, h=540/96)
  # Figure 1c, Version 27.11.2025: export @ Width 280 x Height 540
  
  # --- STEP 5: Supplementary Table of all 24 parameters --
  # including: method of statistics, adj. p-value (FDR)
  # descriptive statistics
  
  # descriptive stats per group & parameter using existing non_normal_params
  # helper function
  format_descr <- function(x, non_normal = FALSE) {
    
    if (!non_normal) {
      m  <- mean(x, na.rm = TRUE)
      sem <- sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
      return(sprintf("%.2f ± %.2f", m, sem))
      
    } else {
      med <- median(x, na.rm = TRUE)
      q1  <- quantile(x, 0.25, na.rm = TRUE)
      q3  <- quantile(x, 0.75, na.rm = TRUE)
      return(sprintf("%.2f (%.2f - %.2f)", med, q1, q3))
    }
  }
  
  table_descr <- df_long %>%
    group_by(Parameter, ident) %>%
    summarise(
      descr = format_descr(
        Value,
        non_normal = unique(Parameter) %in% non_normal_params
      ),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from = ident,
      values_from = descr
    ) %>%
    arrange(Parameter)
  
  # Pairwise statistics table for supplement in paper
  supp_table <- final_results %>%
    left_join(table_descr, by = "Parameter") %>%
    mutate(
      Parameter = param_dict[Parameter],
      `FDR p.adj` = scales::pvalue(p.adj.fdr),
      Significance = p.adj.signif.fdr,
      `Cohen's d` = round(effsize, 3)
    ) %>%
    select(Parameter, `DCF+`, `WFA+`, `Cohen's d`,method, `FDR p.adj`, Significance)

  write_csv(supp_table, file.path(out_dir, "Table_S1_ephys_comparisons.csv"))
} else {
  cat("\n=== MANOVA is NOT significant (p >= 0.05) ===")
  cat("\nNo evidence for overall multivariate difference between WFA+ and DCF+")
  cat("\nUnivariate tests not recommended.\n")
}
```

\large
# 2: Assessing dimensionality of the parameter space
\small
## 2.1: PCA (Principal component analysis)
1. PCA was performed on the z-scored dataset.
2. Variance explained by each principal component was computed.
3. The Kaiser criterion was applied (eigenvalues > 1).
4. A cumulative scree plot was generated (**Fig. 2A**).
5. Variable contributions to the first six principal components were visualized as bar plots (**Fig. S2B**).
\tiny
```{r}
# ============================================================
# PCA ANALYSIS
# ============================================================

# --- Run PCA (scaled variables) ---
pca_results <- prcomp(df_dbl, scale. = TRUE)

# Flip signs for interpretability consistency
pca_results$rotation <- -1 * pca_results$rotation
pca_results$x <- -1 * pca_results$x

# --- Variance explained ---
var_explained <- pca_results$sdev^2 / sum(pca_results$sdev^2)
cum_var_explained <- cumsum(var_explained)

scree <- data.frame(
  PC = seq_along(var_explained),
  Variance = var_explained,
  CumulativeVariance = cum_var_explained
)

# --- Eigenvalues (Kaiser criterion) ---
eigenvalues <- pca_results$sdev^2

pca_df <- data.frame(
  PC = seq_along(var_explained),
  Variance = var_explained,
  CumulativeVariance = cum_var_explained,
  EigenValue = eigenvalues
)

kaiser_components <- eigenvalues[eigenvalues > 1]

# ============================================================
# Scree plot (cumulative variance)
# ============================================================

p_scree <- ggplot(scree, aes(x = PC, y = CumulativeVariance)) +
  geom_col(fill = "grey70", width = 0.8) +
  geom_hline(yintercept = 0.8,
             linetype = "dashed",
             linewidth = 0.3) +
  scale_y_continuous(limits = c(0, 1),
                     breaks = seq(0, 1, by = 0.2)) +
  scale_x_continuous(limits = c(1, 20),
                     breaks = seq(0, 20, by = 5)) +
  labs(x = "Principal Component",
       y = "Cumulative Variance Explained") +
  theme_classic2() +
  theme(
    axis.text = element_text(size = 6, color = "black"),
    axis.title = element_text(size = 8, color = "black"),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )

print(p_scree)

save_fig(p_scree,
         "Fig_2A_Cumulative_Scree_Plot",
         w = 125/96,
         h = 220/96)

# ============================================================
# PCA loadings / variable contributions
# ============================================================

library(FactoMineR)
library(factoextra)
library(gridExtra)

# Contributions per variable
var_contrib <- get_pca_var(pca_results)$contrib

# --- Plot top contributors per PC ---
plots <- list()

for (i in 1:6) {

  plots[[i]] <- fviz_contrib(
    pca_results,
    choice = "var",
    axes = i,
    top = 10
  ) +
    ggtitle(paste("PC", i)) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 8),
      panel.grid = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(color = "black"),
      axis.text = element_text(size = 6, color = "black"),
      axis.title = element_text(size = 8, color = "black"),
      axis.ticks = element_line(linewidth = 0.3)
    )
}

# Arrange panels
p_pca_loadings <- grid.arrange(grobs = plots, nrow = 2, ncol = 3)

# Save figure
ggsave(
  filename = file.path(out_dir, "Suppl_Fig_2B_PCA_loadings.eps"),
  plot = p_pca_loadings,
  width = 500/96,
  height = 350/96,
  units = "in",
  device = "eps"
)

# ============================================================
# Export PCA summary table
# ============================================================

write_csv(
  pca_df,
  file.path(out_dir, "Suppl_Fig_2A_PCA_results.csv")
)
```

\small
## 2.2 Linear Discriminant Analysis (LDA): DCF vs. WFA
- Linear discriminant analysis (LDA) was performed to evaluate separability between WFA+ and DCF+ cells based on electrophysiological parameters.
- All parameters were z-scored prior to analysis to ensure comparability across features.
- An LDA model was fitted using group identity as the response variable.
- Classification performance was assessed using the confusion matrix.
- Accuracy, sensitivity, and positive predictive values were computed for both groups.
- Model generalization performance was evaluated using leave-one-out cross-validation.
- The first linear discriminant (LD1) was used for visualization of group separation.
- LDA coefficients were extracted to identify parameters contributing most strongly to group discrimination.
- Results were visualized as LD1 density plots and coefficient bar plots (**Fig. S1A–B**).

```{r} 
# ========================================================================
# Linear Discriminant Analysis (WFA+ vs DCF+)
# ========================================================================

library(MASS)

# --- Prepare dataset -----------------------------------------------------
# Input: df_int (24 electrophysiological parameters + metadata)

df_lda <- df_int[, !(names(df_int) %in% c("cell", "put_type"))]
df_lda$ident <- as.factor(df_lda$ident)

# --- Standardization -----------------------------------------------------
# Z-score scaling ensures comparability across features

params_lda <- df_lda[, setdiff(names(df_lda), "ident")]
params_scaled <- scale(params_lda)

df_lda_scaled <- data.frame(params_scaled, group = df_lda$ident)

# --- Fit LDA model -------------------------------------------------------
lda_fit <- lda(group ~ ., data = df_lda_scaled)

# Predict class membership and LD scores
lda_pred <- predict(lda_fit)

# LD1 scores for visualization
df_lda_scaled$LD1 <- lda_pred$x[, 1]

# ========================================================================
# CLASSIFICATION PERFORMANCE
# ========================================================================

predicted_classes <- lda_pred$class
actual_classes <- df_lda_scaled$group

# Confusion matrix
conf_matrix <- table(Actual = actual_classes, Predicted = predicted_classes)
print(conf_matrix)

# Accuracy
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)

# Group-wise sensitivity
sensitivity_dcf <- conf_matrix[1,1] / sum(conf_matrix[1,])
sensitivity_wfa <- conf_matrix[2,2] / sum(conf_matrix[2,])

# Positive predictive values
ppv_dcf <- conf_matrix[1,1] / sum(conf_matrix[,1])
ppv_wfa <- conf_matrix[2,2] / sum(conf_matrix[,2])

cat("\n=== LDA Classification Performance ===\n")
cat("Accuracy:", round(accuracy * 100, 1), "%\n")
cat("DCF+ Sensitivity:", round(sensitivity_dcf * 100, 1), "%\n")
cat("WFA+ Sensitivity:", round(sensitivity_wfa * 100, 1), "%\n")
cat("DCF+ PPV:", round(ppv_dcf * 100, 1), "%\n")
cat("WFA+ PPV:", round(ppv_wfa * 100, 1), "%\n")

# ========================================================================
# CROSS-VALIDATION (leave-one-out)
# ========================================================================

lda_cv <- lda(group ~ ., data = df_lda_scaled, CV = TRUE)
cv_accuracy <- mean(lda_cv$class == df_lda_scaled$group)

cat("LOOCV Accuracy:", round(cv_accuracy * 100, 1), "%\n")

# ========================================================================
# VISUALIZATION: LD1 DISTRIBUTION
# ========================================================================

p_lda_density <- ggplot(df_lda_scaled, aes(x = LD1, fill = group)) +
  geom_density(alpha = 0.6, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = my_colors, labels = c("DCF+", "WFA+")) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  labs(title = "LDA Projection", x = "LD1 Score", y = "Density") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 8),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 6),
    axis.line = element_line(color = "black", linewidth = 0.3),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    axis.text = element_text(color = "black", size = 6),
    axis.title = element_text(size = 6),
    panel.grid = element_blank()
  )

print(p_lda_density)
save_fig(p_lda_density,
         "Suppl_Fig_1A_LDA_Projection_DCFvsWFA",
         w = 220/96, h = 250/96,
         device = "pdf") # EPS cannot display transparency

# ========================================================================
# LDA COEFFICIENTS (feature importance)
# ========================================================================

coef_df <- data.frame(
  Parameter = rownames(lda_fit$scaling),
  Coefficient = lda_fit$scaling[, 1]
) %>%
  mutate(Parameter_label = param_dict[Parameter]) %>%
  arrange(desc(abs(Coefficient)))

p_coef <- ggplot(coef_df,
                 aes(x = reorder(Parameter_label, abs(Coefficient)),
                     y = Coefficient)) +
  geom_col(color = "black", fill = "white", width = 0.7) +
  coord_flip() +
  labs(title = "LDA Coefficients (LD1)",
       x = "Parameter", y = "Coefficient") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 8),
    axis.title = element_text(size = 6),
    axis.line = element_line(color = "black", linewidth = 0.3),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    axis.text = element_text(size = 6, color = "black"),
    panel.grid = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.2),
    legend.position = "none"
  )

print(p_coef)
save_fig(p_coef,
         "Suppl_Fig_1B_LDA_Coefficients",
         w = 350/96, h = 280/96)
```

\large
# 3.: Clustering, using different methods.
- 3.1 Unsupervised hierarchircal clustering
- 3.2 Leave-one out Cross validation
- 3.3 LDA of cluster 1 vs cluster 3, plotting the 10 most important discriminators
- 3.2 Stacked plots DCF vs WFA per Cluster
- 3.3 Chi-Square test DCF vs WFA -> sampling per Cluster
- 3.4 LDA Cluster 1 vs. Cluster 3? -> biologically relevant?
\small
## 3.1: Hierarchical clustering
- Unsupervised hierarchical clustering was performed to identify structure in the electrophysiological parameter space.
- Pairwise distances were computed using Manhattan distance, and hierarchical clustering was performed using Ward’s linkage (Ward.D2).
- Cluster stability and optimal cluster number were assessed using silhouette analysis (k = 2–6).
- Dendrograms were generated for 2- and 3-cluster solutions and visualized with cluster-specific color coding (**Fig. 2B**).
- A parameter-by-cell heatmap was generated using z-scored electrophysiological features ordered by hierarchical clustering.
- Parameters were additionally reordered according to a biologically informed sequence for interpretability.
- Results of silhouette analysis were reported in **Fig. S2C**.
```{r}
# ========================================================================
# Hierarchical clustering
# ========================================================================

library(cluster)
library(dendextend)
library(pheatmap)
library(scico)
library(scales)
library(grid)

# --- Standardization -----------------------------------------------------
df_scale <- scale(df_dbl)

# --- Distance matrix -----------------------------------------------------
# Manhattan distance for robustness to outliers
d <- dist(df_scale, method = "manhattan")

# Hierarchical clustering (Ward linkage)
hc <- hclust(d, method = "ward.D2")

# ========================================================================
# SILHOUETTE ANALYSIS (k = 2–6)
# ========================================================================

sil_scores <- data.frame(k = 2:6, avg_silhouette = NA)

for (k in 2:6) {

  clust_level <- cutree(hc, k = k)
  sil <- silhouette(clust_level, d)

  sil_scores$avg_silhouette[k - 1] <- mean(sil[, 3])

  # --- plot (interactive view) ---
  plot(sil,
       main = paste0("Silhouette Plot – ", k, " Clusters"),
       border = NA)

  # --- export ---
  png(
    file.path(out_dir,
              paste0("Suppl_Fig_2C_Silhouette_", k, "_clusters.png")),
    width = 800,
    height = 600,
    res = 150
  )

  plot(sil,
       main = paste0("Silhouette Plot – ", k, " Clusters"),
       border = NA)

  dev.off()
}

print(sil_scores)
write_csv(sil_scores,
           file.path(out_dir, "silhouette_scores.csv"))

# ========================================================================
# DENDROGRAM VISUALIZATION
# ========================================================================

hclust_level <- cutree(hc, k = 3)

dend <- as.dendrogram(hc)

# Cluster color mapping (3-cluster solution)
cluster_cols_3 <- unname(my_colors[c("2", "1", "3")])

dend_colored <- color_branches(dend, k = 3, col = cluster_cols_3)
dend_colored <- set(dend_colored, "labels", NULL)

plot(dend_colored)

pdf(file.path(out_dir, "Fig_2B_Cluster_Dendrogram_B.pdf"),
    width = 430/96,
    height = 240/96)

plot(dend_colored, cex = 0.5)

dev.off()

# # Color the branches based on 2 clusters
# hclust_level <- cutree (hc, k=2)
# dend <- as.dendrogram(hc)
# cluster_cols_2 <- unname(my_colors[c("2", "1")])
# dend_colored <- color_branches(dend, k = 2, col=cluster_cols_2)
# dend_colored <- set(dend_colored, "labels", NULL)
# plot(dend_colored)
# hclust_level <- cutree (hc, k=3)

# ========================================================================
# HEATMAP: PARAMETERS × CELLS
# ========================================================================

# --- parameter order (biological interpretability) ----------------------
custom_order <- c(
  "ap_amp", "ap_width", "ap_rise", "ap_decay",
  "ap_upstroke", "ap_downstroke", "ap_UDR",
  "fAHP_amp", "fAHP_time", "ap_thresh",
  "freq_max", "f_i_slope", "rheobase", "onset_lat",
  "isi_first", "isi_avg", "isi_cv", "isi_adapt",
  "amp_cv", "amp_adapt",
  "RMP", "CC_tau", "CC_Rm", "sag_ratio"
)

# --- transpose for heatmap (features × cells) --------------------------
mat_heat <- t(df_scale)

# --- rescale each feature to [0,1] for visualization -------------------
mat_heat_scaled <- t(apply(mat_heat, 1, function(x) rescale(x, to = c(0, 1))))

# --- clustering of columns (cells) --------------------------------------
row_dist <- dist(mat_heat)
row_clust <- hclust(row_dist)

# reorder rows by biological sequence
mat_heat_scaled_ordered <- mat_heat_scaled[custom_order, ]

# --- color palette ------------------------------------------------------
color_palette <- scico(n = 100, palette = "vik")

# --- heatmap ------------------------------------------------------------
heat_grob <- pheatmap(
  mat_heat_scaled_ordered,
  cluster_rows = FALSE,
  cluster_cols = hc,
  treeheight_col = 0,
  treeheight_row = 0,
  show_colnames = FALSE,
  show_rownames = TRUE,
  fontsize_row = 6,
  color = color_palette,
  border_color = NA
)

save_fig(heat_grob,
         "Fig_2B_Heatmap",
         w = 402.2/96,
         h = 180/96)
```

\small
## 3.2: Cluster validation: Leave-one-trial-out cluster validation (LOOCV)
1. Used for results text, output: Table 3 (**Fig. S3B**)
\tiny
```{r}
library(cluster)
library(dplyr)

# =========================
# LOOCV Cluster Validation
# =========================

k <- 3  # number of clusters from main analysis

# Original cluster assignment (reference solution)
original_clusters <- cutree(hc, k = k)

n <- nrow(df_scale)
loo_assignments <- vector("integer", n)

for (i in 1:n) {

  # -------------------------
  # 1. Leave-one-out sample
  # -------------------------
  train_data <- df_scale[-i, ]

  # -------------------------
  # 2. Recompute clustering
  # -------------------------
  d_train <- dist(train_data, method = "manhattan")
  hc_train <- hclust(d_train, method = "ward.D2")
  train_clusters <- cutree(hc_train, k = k)

  # -------------------------
  # 3. Assign left-out sample
  # -------------------------
  left_out_cell <- df_scale[i, , drop = FALSE]
  d_to_train <- apply(train_data, 1, function(x) sum(abs(x - left_out_cell)))
  nearest_neighbor <- which.min(d_to_train)

  loo_assignments[i] <- train_clusters[nearest_neighbor]
}

# =========================
# Evaluation
# =========================

# Agreement vector
loo_agreement <- loo_assignments == original_clusters
cat("LOOCV cluster assignment accuracy:", mean(loo_agreement), "\n")

# Confusion matrix
loo_table <- table(
  Original = original_clusters,
  LOOCV = loo_assignments
)

print(loo_table)

# Overall stability
loo_accuracy <- sum(diag(loo_table)) / sum(loo_table)
cat("LOOCV Stability:", round(loo_accuracy * 100, 1), "%\n")

# Cluster-wise stability
stability <- diag(loo_table) / rowSums(loo_table) * 100
print(round(stability, 1))

# =========================
# Export
# =========================

loo_results <- data.frame(
  Cluster = paste("Cluster", 1:3),
  Size = as.numeric(table(original_clusters)),
  Stability_pct = round(stability, 1),
  LOOCV_accuracy = round(loo_accuracy * 100, 1)
)

write_csv(loo_results,
          file.path(out_dir, "Suppl_Fig_2C_LOOCV_cluster_stability.csv"))

print(loo_results)
```

\small
## 3.3: LDA of cluster 1 (classical FSI) vs. cluster 3 (borderline-FSI)
1. Assign hierarchical cluster labels (k = 3 solution) to df_int.
2. Subset dataset to Cluster 1 and Cluster 3 only.
3. Select all electrophysiological parameters and standardize variables for LDA input.
4. Fit an LDA model to classify Cluster 1 vs Cluster 3 based on electrophysiological features.
5. Evaluate classification performance using confusion matrix and overall accuracy.
6. Extract LDA coefficients to identify the most discriminative parameters and export results as CSV.
7. Visualize group separation along LD1 using density plots.
8. Select the top 10 discriminative parameters based on absolute LDA coefficients and reshape data into long format.
9. Visualize raw distributions of the top 10 parameters using violin plots, boxplots, and jittered data points, stratified by cluster.
```{r}
library(MASS)
library(dplyr)
library(ggplot2)
library(tidyr)

# =========================
# 1. Cluster assignment
# =========================

df_int$hclust <- factor(cutree(hc, k = 3))

df_int <- df_int %>%
  mutate(
    cluster_label = recode(
      as.character(hclust),
      "1" = "classical FSI",
      "2" = "non-FSI",
      "3" = "borderline-FSI"
    ),
    cluster_label = factor(
      cluster_label,
      levels = c("classical FSI", "borderline-FSI", "non-FSI")
    )
  )

# =========================
# 2. Subset: clusters 1 vs 3
# =========================

df_clusters13 <- df_int %>%
  filter(hclust %in% c("1", "3"))

numeric_cols <- names(df_dbl)

df_c13_numeric <- df_clusters13[, c(numeric_cols, "hclust")]
df_c13_numeric$hclust <- factor(df_c13_numeric$hclust)

# =========================
# 3. Scaling + LDA
# =========================

df_c13_scaled <- data.frame(
  scale(df_c13_numeric[, numeric_cols]),
  hclust = df_c13_numeric$hclust
)

df_c13_scaled$hclust <- factor(df_c13_scaled$hclust, levels = c("1", "3"))

lda_c13_fit <- lda(hclust ~ ., data = df_c13_scaled)
lda_c13_pred <- predict(lda_c13_fit)

# =========================
# 4. Classification performance
# =========================

c13_conf_matrix <- table(
  Actual = df_c13_scaled$hclust,
  Predicted = lda_c13_pred$class
)

c13_accuracy <- sum(diag(c13_conf_matrix)) / sum(c13_conf_matrix)

print(c13_conf_matrix)
cat("Clusters 1 vs 3 LDA Accuracy:", round(c13_accuracy * 100, 1), "%\n")

# =========================
# 5. LDA coefficients
# =========================

lda_coef <- lda_c13_fit$scaling[, 1]

coef_df_c13 <- data.frame(
  Parameter = param_dict[names(lda_coef)],
  Coefficient = lda_coef
) %>%
  arrange(desc(abs(Coefficient)))

write_csv(
  coef_df_c13,
  file.path(out_dir, "Suppl_Fig_2F_LDA_clusters13_coefficients.csv")
)

# =========================
# 6. LD1 density plot
# =========================

df_c13_scaled$LD1 <- lda_c13_pred$x[, 1]

p_lda_c13 <- ggplot(df_c13_scaled, aes(x = LD1, fill = hclust)) +
  geom_density(alpha = 0.6, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = my_colors[c("1", "3")]) +
  labs(x = "LD1 Score", y = "Density") +
  theme_classic2() +
  theme(
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 6),
    legend.position = "bottom"
  )

print(p_lda_c13)
save_fig(p_lda_c13, "Suppl_Fig_2E_LDA_Cluster", w = 2, h = 1.4, dev = "pdf")

# =========================
# 7. Top 10 discriminators
# =========================

top10_coef <- head(coef_df_c13, 10)

p_top10 <- ggplot(top10_coef,
                  aes(x = reorder(Parameter, abs(Coefficient)),
                      y = Coefficient)) +
  geom_col(fill = "white", color = "black") +
  coord_flip() +
  labs(x = "Top 10 discriminating parameters", y = "LD1 coefficient") +
  theme_classic(base_size = 12)

print(p_top10)

ggsave(
  file.path(out_dir, "Suppl_Fig_2F_Clusters13_LDA_top10.pdf"),
  width = 2.6, height = 1.7
)

# =========================
# 8. Raw distributions of top 10
# =========================

top10_raw_names <- names(lda_coef)[order(abs(lda_coef), decreasing = TRUE)[1:10]]

df_c13_long <- df_clusters13 %>%
  pivot_longer(cols = all_of(top10_raw_names),
               names_to = "Parameter",
               values_to = "Value") %>%
  mutate(
    hclust = factor(hclust),
    Parameter = factor(Parameter, levels = top10_raw_names)
  )

levels(df_c13_long$Parameter) <- param_dict_safe[top10_raw_names]

p_raw_top10 <- ggplot(df_c13_long,
                      aes(x = hclust, y = Value, fill = hclust)) +
  geom_violin(alpha = 0.25, linewidth = 0.2) +
  geom_boxplot(width = 0.15, alpha = 0.4, linewidth = 0.2) +
  facet_wrap(~Parameter, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = my_colors[c("1", "3")]) +
  theme_classic2() +
  theme(
    axis.text.x = element_blank(),
    strip.text = element_text(size = 6)
  )

print(p_raw_top10)

save_fig(
  p_raw_top10,
  "Suppl_Fig_2G_Clusters_Raw_Top10",
  w = 3.3, h = 3.0, dev = "pdf"
)
```

\small
## 3.4: Composition of hierarchical clusters across identification groups (DCF+ vs WFA+)
- Visualization of cluster distribution across in situ identification groups
1. Compute proportional distribution of hierarchical clusters within each identification group.
2. Generate stacked bar plot showing relative cluster composition for DCF+ vs WFA+.
3. Visualize cluster proportions using normalized counts (position = "fill").
4. Export figure (**Fig. 2B**) for publication.
\tiny
```{r}

# =========================
# Ensure factor structure
# =========================

df_int$ident <- as.factor(df_int$ident)
df_int$hclust <- as.factor(df_int$hclust)
df_int$cluster_label <- as.factor(df_int$cluster_label)

# =========================
# Stacked bar plot
# =========================

fig_2bar <- ggplot(df_int, aes(x = ident, fill = cluster_label)) +
  geom_bar(position = "fill", width = 0.8) +  # relative proportions

  scale_fill_manual(
    values = my_colors,
    guide = guide_legend(
      override.aes = list(size = 1)  # smaller legend symbols
    )
  ) +

  labs(
    x = "In situ identification",
    y = "Relative count",
    fill = "Cluster"
  ) +

  theme_classic() +
  theme(
    axis.text.x  = element_text(size = 6),
    axis.text.y  = element_text(size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 6)
  )

print(fig_2bar)

save_fig(
  fig_2bar,
  "Fig_2B_Stacked_Bar",
  w = 210 / 96,
  h = 220 / 96,
  device = "eps"
)
```

\small
## 3.5: Cluster–Group Association (Chi-square / Fisher’s Exact Test)
- Statistical test of association between identification group and cluster assignment (reported in Results; no figure).
\tiny
```{r}
# =========================
# Cluster–Identification Association
# Chi-square + Fisher’s Exact Test
# =========================

library(dplyr)
library(tidyr)

# -------------------------
# 1. Count observations
# -------------------------

counts <- df_int %>%
  group_by(ident, cluster_label) %>%
  summarise(n = n(), .groups = "drop")

# -------------------------
# 2. Build contingency table
# -------------------------

contingency_tbl <- counts %>%
  pivot_wider(
    names_from = cluster_label,
    values_from = n,
    values_fill = 0
  )

print("Contingency table")
print(contingency_tbl)

# Convert to matrix format
fisher_matrix <- as.matrix(contingency_tbl[, -1])
rownames(fisher_matrix) <- contingency_tbl$ident

# -------------------------
# 3. Fisher's Exact Test
# -------------------------
fisher_res <- fisher.test(fisher_matrix)
print(fisher_res)

# -------------------------
# 4. Chi-square test (for comparison only)
# -------------------------

chi_matrix <- fisher_matrix

chi_result <- chisq.test(chi_matrix)

print("Chi-square test (note: assumptions may be violated for small counts)")
print(chi_result)

# Expected counts (optional inspection)
# chi_result$expected
```

\large
# 4. Uniform Manifold Approximation and Projection (UMAP) of electrophysiological data
\small
1. Compute UMAP embedding from standardized electrophysiological parameter space
2. Add UMAP coordinates to dataframe
3. Merge cluster assignments (hierarchical clustering) and identification labels (DCF+/WFA+)
4. Visualize UMAP embedding using cluster identity as color and experimental identification as shape
5. (Optional) alternative UMAP visualizations are commented out for exploratory analysis

## 4.1 UMAP with clusters as color code and identification method as shape
- **Fig. 2C**
```{r}
# =========================
# UMAP: parameter space embedding
# =========================

set.seed(1234)

# Scale parameter matrix
df_scale <- scale(df_dbl)

# Compute UMAP embedding
umap_fit <- umap(
  df_scale,
  n_neighbors = 50,
  spread = 0.1,
  min_dist = 0.05
)

# =========================
# Build UMAP dataframe
# =========================

umap_df <- umap_fit$layout %>%
  as.data.frame() %>%
  rename(
    UMAP1 = V1,
    UMAP2 = V2
  ) %>%
  mutate(cell = pull(cells, cell)) %>%
  inner_join(df_meta, by = "cell")

# Add clustering information
umap_df <- umap_df %>%
  left_join(
    df_int %>% dplyr::select(cell, hclust, cluster_label),
    by = "cell"
  )

# =========================
# Main UMAP plot (Fig. 2C)
# =========================

fig_2UMAP <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster_label, shape = ident)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = my_colors) +
  scale_shape_manual(values = c("DCF+" = 16, "WFA+" = 1)) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    color = "Cluster",
    shape = "Identification"
  ) +
  theme_classic2() +
  theme(
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 6)
  )

print(fig_2UMAP)

save_fig(
  fig_2UMAP,
  "Fig_2C_UMAP",
  w = 390/96,
  h = 290/96,
  device = "eps"
)

# =========================
# Alternative / exploratory UMAP plots (commented)
# =========================

# umap_df %>%
#   ggplot(aes(x = UMAP1, y = UMAP2, color = ident, shape = put_type)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = my_colors) +
#   scale_fill_manual(values = my_colors) +
#   theme_classic()

# umap_df %>%
#   ggplot(aes(x = UMAP1, y = UMAP2, color = hclust, shape = put_type)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = my_colors) +
#   scale_shape_manual(values = c("PC" = 17, "FSIN" = 16, "Non-FSIN" = 1)) +
#   theme_classic2()

# x_range <- range(umap_df$UMAP1)
# y_range <- range(umap_df$UMAP2)

# umap_df %>%
#   ggplot(aes(x = UMAP1, y = UMAP2, color = hclust)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = my_colors) +
#   theme_classic2() +
#   xlim(x_range) +
#   ylim(y_range) +
#   labs(title = "UMAP of All Groups")

# umap_df_dcf <- umap_df[umap_df$ident == "DCF+", ]

# umap_df_dcf %>%
#   ggplot(aes(x = UMAP1, y = UMAP2, color = hclust)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = my_colors) +
#   theme_classic2() +
#   xlim(x_range) +
#   ylim(y_range) +
#   labs(title = "UMAP of DCF+ Group Only")

# umap_df_wfa <- umap_df[umap_df$ident == "WFA+", ]

# umap_df_wfa %>%
#   ggplot(aes(x = UMAP1, y = UMAP2, color = hclust)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = my_colors) +
#   theme_classic2() +
#   xlim(x_range) +
#   ylim(y_range) +
#   labs(title = "UMAP of WFA+ Group Only")
```

\large 
## 4.2: UMAP exploration - Functional gradients per parameter over UMAP space
- Visualization of electrophysiological parameter distributions across the UMAP embedding
- Each panel shows one parameter as a continuous gradient over UMAP space
- **Fig. 2D**
\tiny
```{r, fig.width = 4, fig.height = 3}
# =========================
# Merge UMAP with electrophysiological parameters
# =========================

umap_params <- left_join(
  umap_df,
  df_original[, c(
    "cell",
    "VC_Rs", "RMP", "CC_tau", "CC_Rm", "sag_ratio", "rheobase",
    "freq_max", "f_i_slope", "onset_lat",
    "isi_first", "isi_avg", "isi_cv", "isi_adapt",
    "amp_cv", "amp_adapt", "burst",
    "ap_amp", "ap_thresh", "ap_width", "ap_rise", "ap_decay",
    "ap_upstroke", "ap_downstroke", "ap_UDR",
    "fAHP_amp", "fAHP_time"
  )],
  by = "cell"
)

# =========================
# Parameter sets
# =========================

umap_custom <- c(
  "ap_width", "ap_upstroke", "fAHP_amp",
  "ap_thresh", "freq_max", "rheobase",
  "amp_adapt", "isi_adapt", "sag_ratio",
  "CC_Rm", "CC_tau", "RMP"
)

umap_full <- c(
  "RMP", "CC_tau", "CC_Rm", "sag_ratio", "rheobase",
  "freq_max", "f_i_slope", "onset_lat",
  "isi_first", "isi_avg", "isi_cv", "isi_adapt",
  "amp_cv", "amp_adapt",
  "ap_amp", "ap_thresh", "ap_width", "ap_rise", "ap_decay",
  "ap_upstroke", "ap_downstroke", "ap_UDR",
  "fAHP_amp", "fAHP_time"
)

# =========================
# Long-format transformation
# =========================

umap_long <- umap_params %>%
  pivot_longer(
    cols = umap_custom,
    names_to = "parameter",
    values_to = "value"
  ) %>%
  group_by(parameter) %>%
  mutate(
    scaled_value = scales::rescale(abs(value))
  ) %>%
  ungroup() %>%
  mutate(
    parameter = factor(parameter, levels = umap_custom)
  )

# =========================
# UMAP gradient plot (Fig. 2D)
# =========================

fig_2gradient <- ggplot(umap_long, aes(x = UMAP1, y = UMAP2, color = scaled_value)) +
  geom_point(size = 0.8) +
  scale_color_scico(palette = "lajolla") +
  facet_wrap(
    ~parameter,
    scales = "free",
    ncol = 3,
    labeller = as_labeller(param_dict)
  ) +
  labs(
    title = NULL,
    color = "Scaled Value"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 8),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 6),
    legend.text = element_text(size = 6)
  ) +
  guides(
    color = guide_colorbar(
      barwidth = 0.5,
      barheight = 4
    )
  )

print(fig_2gradient)

save_fig(
  fig_2gradient,
  "Fig_2D_UMAP_gradients",
  w = 410/96,
  h = 500/96,
  device = "eps"
)
```

\large
# 5 Further exploration of data of single patients
- Map metadata on UMAP
- **Figure S3**
```{r}
library(patchwork)

# =========================
# Helper: common ggplot theme (kept inline for consistency)
# =========================

base_umap_theme <- theme_classic2() +
  theme(
    plot.title = element_text(size = 8, hjust = 0.5),
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 6),
    axis.line = element_line(linewidth = 0.3),
    axis.ticks = element_line(linewidth = 0.3),
    legend.position = "none"
  )

facet_umap_theme <- theme_classic2() +
  theme(
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 6),
    axis.line = element_line(linewidth = 0.3),
    axis.ticks = element_line(linewidth = 0.3),
    legend.position = "none",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6),
    strip.text = element_blank(),
    strip.background = element_blank()
  )

# =========================
# Patient
# =========================

umap_df$pat <- as.factor(umap_df$pat)

umap_pat <- ggplot(umap_df, aes(UMAP1, UMAP2, color = pat)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  labs(title = "UMAP colored by Patient", color = "Patient") +
  base_umap_theme

umap_pat_facet <- ggplot(umap_df, aes(UMAP1, UMAP2, color = pat)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  facet_wrap(~pat, ncol = 6) +
  labs(title = NULL, color = "Patient") +
  facet_umap_theme

fig_umap_pat <- (umap_pat | umap_pat_facet) +
  plot_layout(widths = c(4, 6), axes = "collect_x", axis_titles = "collect_x")

print(fig_umap_pat)
save_fig(fig_umap_pat, "Suppl_Fig_3A_pat", w = 550/96, h = 200/96, device = "pdf")

# =========================
# Sex
# =========================

umap_df$sex <- as.factor(umap_df$sex)

umap_sex <- ggplot(umap_df, aes(UMAP1, UMAP2, color = sex)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  labs(title = "UMAP colored by patient sex", color = "Sex") +
  base_umap_theme

umap_sex_facet <- ggplot(umap_df, aes(UMAP1, UMAP2, color = sex)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  facet_wrap(~sex) +
  labs(title = NULL, color = "Sex") +
  facet_umap_theme

fig_umap_sex <- (umap_sex | umap_sex_facet) +
  plot_layout(widths = c(4, 6), axes = "collect_x", axis_titles = "collect_x")

print(fig_umap_sex)
save_fig(fig_umap_sex, "Suppl_Fig_3B_sex", w = 550/96, h = 200/96, device = "pdf")

# =========================
# Clinic
# =========================

umap_df$clinic <- as.factor(umap_df$clinic)

umap_clin <- ggplot(umap_df, aes(UMAP1, UMAP2, color = clinic)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  labs(title = "UMAP colored by location of clinic", color = "Clinic") +
  base_umap_theme

umap_clin_facet <- ggplot(umap_df, aes(UMAP1, UMAP2, color = clinic)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  facet_wrap(~clinic) +
  labs(title = NULL, color = "Clinic") +
  facet_umap_theme

fig_umap_clin <- (umap_clin | umap_clin_facet) +
  plot_layout(widths = c(4, 6), axes = "collect_x", axis_titles = "collect_x")

print(fig_umap_clin)
save_fig(fig_umap_clin, "Suppl_Fig_3C_clin", w = 550/96, h = 200/96, device = "pdf")

# =========================
# Area
# =========================

umap_df$area <- as.factor(umap_df$area)

umap_area <- ggplot(umap_df, aes(UMAP1, UMAP2, color = area)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  labs(title = "UMAP colored by resection area", color = "Resection area") +
  base_umap_theme

umap_area_facet <- ggplot(umap_df, aes(UMAP1, UMAP2, color = area)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  facet_wrap(~area) +
  labs(title = NULL, color = "Area") +
  facet_umap_theme

fig_umap_area <- (umap_area | umap_area_facet) +
  plot_layout(widths = c(4, 6), axes = "collect_x", axis_titles = "collect_x")

print(fig_umap_area)
save_fig(fig_umap_area, "Suppl_Fig_3D_area", w = 550/96, h = 200/96, device = "pdf")

# =========================
# Pathology
# =========================

umap_df$patho <- as.factor(umap_df$patho)

umap_patho <- ggplot(umap_df, aes(UMAP1, UMAP2, color = patho)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  labs(title = "UMAP colored by operation type ('patho')", color = "Pathology") +
  base_umap_theme

umap_patho_facet <- ggplot(umap_df, aes(UMAP1, UMAP2, color = patho)) +
  geom_point(size = 0.6) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  facet_wrap(~patho) +
  labs(title = NULL, color = "Pathology") +
  facet_umap_theme

fig_umap_patho <- (umap_patho | umap_patho_facet) +
  plot_layout(widths = c(4, 6), axes = "collect_x", axis_titles = "collect_x")

print(fig_umap_patho)
save_fig(fig_umap_patho, "Suppl_Fig_3D_patho", w = 550/96, h = 200/96, device = "pdf")

# =========================
# Age
# =========================

umap_age <- ggplot(umap_df, aes(UMAP1, UMAP2, color = age)) +
  geom_point(size = 0.6) +
  scale_color_viridis_c(option = "D") +
  labs(title = "UMAP colored by age", color = "Age") +
  base_umap_theme +
  theme(
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6)
  )

print(umap_age)
save_fig(umap_age, "Suppl_Fig_3E_age", w = 290/96, h = 185/96, device = "pdf")

# =========================
# Putative type
# =========================

umap_df$put_type <- as.factor(umap_df$put_type)

ggplot(umap_df, aes(UMAP1, UMAP2, color = put_type)) +
  geom_point(size = 2) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_classic2() +
  labs(title = "UMAP colored by experimenter rating", color = "Putative cell type")

ggplot(umap_df, aes(UMAP1, UMAP2, color = put_type)) +
  geom_point(size = 2) +
  facet_wrap(~put_type) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_classic2() +
  labs(title = "UMAP Plots by Putative Type", color = "Putative Type")

# =========================
# Resection area (redundant exploratory plots)
# =========================

ggplot(umap_df, aes(UMAP1, UMAP2, color = area)) +
  geom_point(size = 2) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_classic2() +
  labs(title = "UMAP colored by resection area", color = "Resection area")

ggplot(umap_df, aes(UMAP1, UMAP2, color = area)) +
  geom_point(size = 2) +
  facet_wrap(~area) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_classic2() +
  labs(title = "UMAP Plots by Resection Area", color = "Area")

# =========================
# Highlight specific cells
# =========================

ggplot(umap_df, aes(UMAP1, UMAP2, color = cluster_label, shape = ident)) +
  geom_point(size = 1.5, alpha = 0.3) +
  geom_point(
    data = umap_df %>% filter(cell %in% c(
      "E437-2",
      "E384-4",
      "E388-3"
    )),
    aes(UMAP1, UMAP2, color = cluster_label),
    size = 3,
    alpha = 1
  ) +
  scale_color_manual(values = my_colors) +
  scale_shape_manual(values = c("DCF+" = 16, "WFA+" = 1)) +
  labs(x = "UMAP1", y = "UMAP2", color = "Cluster", shape = "Identification") +
  theme_classic2() +
  theme(
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6)
  )
```

\large
# 6 Exploration of anatomical markers (PV and SST) within electrophysiology (Figure 3)
- Extend df_anat with immunohistochemistry-based costaining labels (PV, SST), including clear, potentially positive/negative, and ambiguous categories
- Merge anatomical data with UMAP coordinates, clustering results, metadata, and electrophysiological parameters via cell ID
- Generate summary tables of costaining distributions across groups
- Subset PV+ and SST+ populations for downstream analyses
1. Define groups based on immunohistochemical labeling (PV or SST)
2. Visualize PV+ and SST+ cells in UMAP space, highlighting them on the global embedding (**Fig. 3A–C**)
3. Compare electrophysiological properties between DCF+ and WFA+ within PV+ cells using parameter-wise analyses (violin/box plots)
4. Perform Fisher’s exact tests (and descriptive chi-square comparisons) to assess associations between cell identity (DCF/WFA) and PV/SST positivity, reporting effect sizes and p-values
```{r}
# ------------------------------
# --- Add costaining labels ---
# ------------------------------

df_anat <- df_anat_raw %>%
  mutate(
    anat_dist = anat_dist * 10^-3,

    anat_costain = case_when(
      is.na(anat_morph) ~ "no rec",
      anat_pv == "+" & anat_sst == "+" ~ "pv+sst+",
      anat_pv == "+" & anat_sst == "-" ~ "pv+",
      anat_sst == "+" & anat_pv == "-" ~ "sst+",
      anat_pv == "?" & anat_sst == "-" ~ "potentially pv+",
      anat_sst == "?" & anat_pv == "-" ~ "potentially sst+",
      anat_pv == "-" & anat_sst == "-" ~ "both costains negative",
      TRUE ~ "unclear"
    ),

    anat_pv = case_when(
      is.na(anat_morph) ~ "no rec",
      anat_pv == "+" ~ "pv_pos",
      anat_pv == "-" ~ "pv_neg",
      anat_pv == "?" ~ "unclear"
    ),

    anat_sst = case_when(
      is.na(anat_morph) ~ "no rec",
      anat_sst == "+" ~ "sst_pos",
      anat_sst == "-" ~ "sst_neg",
      anat_sst == "?" ~ "unclear"
    )
  )

# ------------------------------
# --- Merge anatomical + UMAP + electrophysiology ---
# ------------------------------

df_anat <- umap_df %>%
  left_join(
    df_anat %>%
      dplyr::select(cell, anat_morph, anat_costain, anat_pv, anat_sst, anat_dist),
    by = "cell"
  ) %>%
  left_join(
    df_int %>%
      dplyr::select(-ident, -put_type, -hclust, -cluster_label),
    by = "cell"
  )

# Convert to factors
df_anat <- df_anat %>%
  mutate(
    anat_costain = factor(anat_costain),
    anat_pv      = factor(anat_pv),
    anat_sst     = factor(anat_sst),
    anat_morph   = factor(anat_morph)
  )

# ------------------------------
# --- Descriptive statistics ---
# ------------------------------

cat("PV costaining\n")
table(df_anat$ident, df_anat$anat_pv)

cat("\nSST costaining\n")
table(df_anat$ident, df_anat$anat_sst)

cat("\nMorphology\n")
table(df_anat$ident, df_anat$anat_morph)

table(df_anat$anat_morph, df_anat$anat_pv)
table(df_anat$anat_morph, df_anat$anat_sst)

# ------------------------------
# --- Primary Fisher analysis (PV only, no unclear) ---
# ------------------------------

df_primary <- df_anat %>%
  filter(anat_pv %in% c("pv_pos", "pv_neg")) %>%
  droplevels()

cont_table  <- table(df_primary$ident, df_primary$anat_pv)
cont_table2 <- table(df_primary$cluster_label, df_primary$anat_pv)

cont_table
fisher.test(cont_table)

cont_table2
fisher.test(cont_table2)

# ------------------------------
# --- Summary costaining distribution ---
# ------------------------------

summary_table <- table(df_anat$anat_costain)
summary_table

# ------------------------------
# --- Counts of PV+ cells ---
# ------------------------------

n_dcf_pv <- df_anat %>%
  filter(ident == "DCF+", anat_costain == "pv+") %>%
  nrow()

n_wfa_pv <- df_anat %>%
  filter(ident == "WFA+", anat_costain == "pv+") %>%
  nrow()

cat("Number of DCF+ PV+ cells:", n_dcf_pv, "\n")
cat("Number of WFA+ PV+ cells:", n_wfa_pv, "\n")

# ------------------------------
# --- Additional contingency tests ---
# ------------------------------

fisher_mat <- matrix(
  c(20, 31,   # WFA
    12, 52),  # DCF
  nrow = 2,
  byrow = TRUE
)

fisher.test(fisher_mat)

table(df_anat$ident, df_anat$anat_costain, useNA = "ifany")
table(df_anat$anat_morph, df_anat$cluster_label, useNA = "ifany")
table(df_anat$anat_morph, df_anat$anat_costain, useNA = "ifany")
table(df_anat$anat_morph, df_anat$ident, useNA = "ifany")

# ------------------------------
# --- Summary statistics per costain group ---
# ------------------------------

table_anat_groups <- df_anat %>%
  group_by(anat_costain) %>%
  summarise(
    across(
      all_of(c(
        "RMP", "CC_Rm", "CC_tau", "sag_ratio", "freq_max", "f_i_slope",
        "isi_adapt", "amp_adapt", "ap_thresh", "ap_width", "ap_rise",
        "ap_decay", "fAHP_amp"
      )),
      list(
        n    = ~sum(!is.na(.)),
        mean = ~mean(., na.rm = TRUE),
        sd   = ~sd(., na.rm = TRUE),
        med  = ~median(., na.rm = TRUE),
        min  = ~min(., na.rm = TRUE),
        max  = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

table_anat_groups

# --------------------------------------------
# --- Figure 3C: UMAP with PV+ highlighted ---
# --------------------------------------------

df_anat_pv <- df_anat %>%
  filter(anat_pv == "pv_pos")

fig_3PV <- ggplot(df_anat, aes(x = UMAP1, y = UMAP2)) +
  # background cells
  geom_point(
    aes(shape = ident, color = cluster_label),
    size = 0.9,
    alpha = 0.1
  ) +
  # PV+ highlighted
  geom_point(
    data = df_anat_pv,
    aes(shape = ident, color = cluster_label),
    size = 1.5,
    alpha = 1
  ) +
  scale_shape_manual(values = c("DCF+" = 16, "WFA+" = 1), name = "Identification") +
  scale_color_manual(values = my_colors, name = "Cluster") +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_classic2() +
  theme(
    axis.text.x  = element_text(size = 6),
    axis.text.y  = element_text(size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 6),
    axis.line    = element_line(size = 0.3),
    axis.ticks   = element_line(size = 0.3)
  )

print(fig_3PV)
save_fig(fig_3PV, "Fig_3C_UMAP_PV", w = 275/96, h = 150/96, device = "pdf")


# --------------------------------------------
# --- Figure: UMAP with SST+ highlighted ---
# --------------------------------------------

df_anat_sst <- df_anat %>%
  filter(anat_sst == "sst_pos")

fig_3sst <- ggplot(df_anat, aes(x = UMAP1, y = UMAP2)) +
  geom_point(
    aes(shape = ident, color = cluster_label),
    size = 0.9,
    alpha = 0.1
  ) +
  geom_point(
    data = df_anat_sst,
    aes(shape = ident, color = cluster_label),
    size = 1.5,
    alpha = 1
  ) +
  scale_shape_manual(values = c("DCF+" = 16, "WFA+" = 1), name = "Identification") +
  scale_color_manual(values = my_colors, name = "Cluster") +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_classic2() +
  theme(
    axis.text.x  = element_text(size = 6),
    axis.text.y  = element_text(size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 6),
    axis.line    = element_line(size = 0.3),
    axis.ticks   = element_line(size = 0.3)
  )

print(fig_3sst)

# --------------------------------------------
# --- UMAP: morphology overlay ---
# --------------------------------------------

ggplot(df_anat, aes(x = UMAP1, y = UMAP2)) +
  geom_point(
    aes(color = anat_morph, shape = ident),
    size = 1.2,
    alpha = 1
  ) +
  scale_shape_manual(values = c("DCF+" = 16, "WFA+" = 1), name = "Identification") +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_classic2() +
  theme(
    axis.text.x  = element_text(size = 6),
    axis.text.y  = element_text(size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 6)
  )

# -------------------------------------------------
# --- Figure 3D: pairwise comparisons, PV+ only ---
# -------------------------------------------------

# ------------------------------
# --- Reshape data ---
# ------------------------------

df_anat_pv_long <- df_anat_pv %>%
  pivot_longer(
    cols = all_of(param_names),
    names_to = "Parameter",
    values_to = "Value"
  ) %>%
  filter(!is.na(Parameter), Parameter %in% custom_params)

# ------------------------------
# --- Statistical testing ---
# ------------------------------

p_values_pv <- df_anat_pv_long %>%
  group_by(Parameter) %>%
  nest() %>%
  mutate(
    # Shapiro normality per group
    normality = map(data, ~ .x %>%
      group_by(ident) %>%
      summarise(
        p_shapiro = shapiro.test(Value)$p.value,
        .groups = "drop"
      )
    ),

    # normality decision
    is_normal = map_lgl(normality, ~ all(.x$p_shapiro > 0.05)),

    # appropriate test
    p_value = map2_dbl(data, is_normal, ~ {
      if (.y) {
        t.test(Value ~ ident, data = .x)$p.value
      } else {
        wilcox.test(Value ~ ident, data = .x, exact = FALSE)$p.value
      }
    }),

    test_used = if_else(is_normal, "t-test", "wilcoxon")
  ) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p_value, method = "fdr"),
    sig_label = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  ) %>%
  select(Parameter, test_used, p_value, p_adj, sig_label)

print(p_values_pv)

# ------------------------------
# --- Annotation positions ---
# ------------------------------

y_positions <- df_anat_pv_long %>%
  group_by(Parameter) %>%
  summarise(
    y.position = max(Value, na.rm = TRUE) +
      0.15 * abs(max(Value, na.rm = TRUE)),
    .groups = "drop"
  )

stat_anno <- p_values_pv %>%
  left_join(y_positions, by = "Parameter") %>%
  mutate(
    group1 = "DCF+",
    group2 = "WFA+",
    label = case_when(
      p_adj < 0.05 ~ sig_label,
      p_adj < 0.1  ~ sprintf("p = %.3f", p_adj),
      TRUE ~ "ns"
    )
  )

# ------------------------------
# --- Plot function template ---
# ------------------------------

plot_pv_params <- function(data, anno, params, out_name, height) {

  data_sel <- filter(data, Parameter %in% params)
  anno_sel <- filter(anno, Parameter %in% params)

  ggplot(data_sel, aes(x = ident, y = Value, color = ident)) +
    coord_cartesian(clip = "off") +
    geom_jitter(width = 0.1, size = 0.6, alpha = 1, na.rm = TRUE) +
    geom_violin(width = 0.7, trim = TRUE, alpha = 0.25, linewidth = 0.3, na.rm = TRUE) +
    geom_boxplot(width = 0.15, outlier.shape = NA, color = "black",
                 alpha = 0.3, size = 0.3, na.rm = TRUE) +
    stat_pvalue_manual(
      anno_sel,
      label = "label",
      y.position = anno_sel$y.position,
      tip.length = 0.01,
      size = 2,
      bracket.size = 0.3,
      hide.ns = TRUE
    ) +
    scale_color_manual(values = my_colors) +
    facet_wrap(
      ~factor(Parameter, levels = params),
      scales = "free_y",
      ncol = 3,
      labeller = as_labeller(param_dict),
      drop = FALSE,
      as.table = TRUE
    ) +
    labs(title = NULL, x = NULL, y = NULL) +
    theme_classic(base_size = 14) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 6),
      axis.text.y = element_text(size = 6, color = "black"),
      axis.line = element_line(color = "black", linewidth = 0.3),
      axis.ticks = element_line(linewidth = 0.3),
      strip.background = element_blank(),
      strip.text = element_text(size = 8, face = "plain", margin = margin(b = 10)),
      legend.position = "none",
      plot.margin = margin(t = 15, r = 5, b = 5, l = 5)
    )
}

# ------------------------------
# --- Full PV panel ---
# ------------------------------

fig_params_pv <- plot_pv_params(
  df_anat_pv_long,
  stat_anno,
  custom_params,
  "Suppl_Fig_5_all_params_PV",
  height = 540/96
)

print(fig_params_pv)
save_fig(fig_params_pv, "Suppl_Fig_5_all_params_PV", w = 280/96, h = 540/96)

# ------------------------------
# --- Selected parameters ---
# ------------------------------

selected_params <- c("CC_tau", "ap_width", "freq_max")

fig_params_pv_sel <- plot_pv_params(
  df_anat_pv_long,
  stat_anno,
  selected_params,
  "Fig_3D_params_PV",
  height = 180/96
)

print(fig_params_pv_sel)
save_fig(fig_params_pv_sel, "Fig_3D_params_PV", w = 280/96, h = 180/96) 
```

\large
# 6.1 Analysis and plotting of depth information
- import and merge IMARIS-derived depth datasets from three human patient samples, standardize column names, and combine into a unified dataframe
- assign patient identity as factor and extract cortical depth (distance from pia) for downstream analysis
- perform descriptive statistics and test depth normality (Shapiro–Wilk)
- compare depth between DCF+ and WFA+ cells using t-test and summarize group-wise SEM statistics
- test depth differences across clusters using ANOVA with Tukey post hoc comparisons
- visualize UMAP with depth as continuous feature (grey background for all cells, depth-colored overlay for reconstructed cells, PV+ cells highlighted)
- plot depth distributions for identification groups and clusters using jitter + boxplots (with optional violin overlays)
- reverse y-axis to reflect cortical depth organization (pia → deep layers)
- integrate IMARIS layer annotations (L2/3, L4, L5/6) to visualize layer-specific depth distributions across patients using faceted plots
- provide alternative visual encodings (color-coded layers vs patient-coded distributions) to assess inter-individual variability
```{r}
library(gghalves)
library(patchwork)
library(janitor)
library(rstatix)
library(readr)
library(dplyr)
library(ggplot2)

# ------------------
# --- Load IMARIS depth data ---
# ------------------

df_pat1 <- read_csv(file.path("data", "WFA_HUMAN_QUANT_IMARIS_PatA.csv")) %>%
  mutate(pat = "Patient A")

df_pat2 <- read_csv(file.path("data", "WFA_HUMAN_QUANT_IMARIS_PatB.csv")) %>%
  mutate(pat = "Patient B")

df_pat3 <- read_csv(file.path("data", "WFA_HUMAN_QUANT_IMARIS_PatC.csv")) %>%
  mutate(pat = "Patient C")

highlight_colors <- c(
  "Patient A" = "#A6CEE3",
  "Patient B" = "#F4A3C4",
  "Patient C" = "#FDBF6F"
)

df_depth <- bind_rows(df_pat1, df_pat2, df_pat3) %>%
  rename(depth = `Distance from pia (µm)`) %>%
  clean_names() %>%
  rename(net_volume = net_volume_mm3) %>%
  mutate(pat = factor(pat, levels = names(highlight_colors)))

max_depth <- max(df_depth$depth, na.rm = TRUE)

# ------------------
# --- Descriptive stats ---
# ------------------

shapiro.test(df_anat$anat_dist)

mean_anat <- mean(df_anat$anat_dist, na.rm = TRUE)
sem_anat  <- sd(df_anat$anat_dist, na.rm = TRUE) /
  sqrt(sum(!is.na(df_anat$anat_dist)))

sprintf("%.2f ± %.2f", mean_anat, sem_anat)

t.test(anat_dist ~ ident, data = df_anat)

sem_table <- df_anat %>%
  filter(ident %in% c("DCF+", "WFA+")) %>%
  group_by(ident) %>%
  summarise(
    n = sum(!is.na(anat_dist)),
    mean = mean(anat_dist, na.rm = TRUE),
    sd = sd(anat_dist, na.rm = TRUE),
    sem = sd / sqrt(n),
    .groups = "drop"
  )

anova_model <- aov(anat_dist ~ cluster_label, data = df_anat)
summary(anova_model)
TukeyHSD(anova_model)

sem_table2 <- df_anat %>%
  group_by(cluster_label) %>%
  summarise(
    n = sum(!is.na(anat_dist)),
    mean = mean(anat_dist, na.rm = TRUE),
    sd = sd(anat_dist, na.rm = TRUE),
    sem = sd / sqrt(n),
    .groups = "drop"
  )

# ------------------
# --- Figure 3B: UMAP depth ---
# ------------------

fig_3depth <- ggplot(df_anat, aes(UMAP1, UMAP2)) +
  geom_point(color = "grey50", size = 0.4, alpha = 0.15) +
  geom_point(
    data = subset(df_anat, !is.na(anat_dist)),
    aes(color = anat_dist),
    size = 0.6, alpha = 0.9
  ) +
  geom_point(
    data = df_anat_pv,
    aes(color = anat_dist),
    size = 0.6, alpha = 1
  ) +
  scale_color_scico(name = "distance from pia (µm)", palette = "lajolla") +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_classic2() +
  theme(
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 6),
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6),
    axis.line = element_line(size = 0.3),
    axis.ticks = element_line(size = 0.3)
  )

print(fig_3depth)
save_fig(fig_3depth, "Fig_3B_umap_depth", w = 250/96, h = 180/96, device = "pdf")

# ------------------
# --- Depth distributions (Fig 3A) ---
# ------------------

p1_wfa_dcf <- df_anat %>%
  filter(!is.na(anat_dist)) %>%
  ggplot(aes(ident, anat_dist, color = ident)) +
  geom_jitter(width = 0.15, size = 0.4, alpha = 0.7) +
  geom_boxplot(width = 0.10, alpha = 0.3, size = 0.3,
               position = position_nudge(x = -0.25),
               outlier.shape = NA) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_y_reverse(limits = c(max_depth, 0)) +
  scale_color_manual(values = my_colors) +
  labs(x = "in situ identification", y = "distance from pia (µm)") +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 6)
  )

p2_clusters <- df_anat %>%
  filter(!is.na(anat_dist)) %>%
  ggplot(aes(cluster_label, anat_dist, color = cluster_label)) +
  geom_jitter(width = 0.15, size = 0.4, alpha = 0.7) +
  geom_boxplot(width = 0.10, alpha = 0.3, size = 0.3,
               position = position_nudge(x = -0.25),
               outlier.shape = NA) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_y_reverse(limits = c(max_depth, 0)) +
  scale_color_manual(values = my_colors[c("classical FSI","borderline-FSI","non-FSI")]) +
  labs(x = "cluster", y = NULL) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig3_cell_depth_v1 <- (p1_wfa_dcf | p2_clusters) +
  plot_layout(widths = c(4, 5), axes = "collect_x", axis_titles = "collect_x")

print(fig3_cell_depth_v1)
save_fig(fig3_cell_depth_v1, "Fig_3A_cell_depth_v1", w = 350/96, h = 200/96, device = "pdf")

# ------------------
# --- Version 2 (greyscale variant) ---
# ------------------

p3_wfa_dcf <- df_anat %>%
  filter(!is.na(anat_dist)) %>%
  ggplot(aes(ident, anat_dist)) +
  geom_jitter(width = 0.15, size = 0.4, color = "gray50") +
  geom_boxplot(width = 0.1, alpha = 0.3, size = 0.3,
               outlier.shape = NA,
               position = position_nudge(x = -0.25)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_y_reverse(limits = c(max_depth, 0)) +
  labs(x = "in situ identification", y = "distance from pia (µm)") +
  theme_classic(base_size = 12)

p4_clusters <- df_anat %>%
  filter(!is.na(anat_dist)) %>%
  ggplot(aes(cluster_label, anat_dist)) +
  geom_jitter(width = 0.15, size = 0.4, color = "gray50") +
  geom_boxplot(width = 0.1, alpha = 0.3, size = 0.3,
               outlier.shape = NA,
               position = position_nudge(x = -0.25)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_y_reverse(limits = c(max_depth, 0)) +
  labs(x = "cluster", y = NULL) +
  theme_classic(base_size = 12)

fig3_cell_depth_v2 <- (p3_wfa_dcf | p4_clusters) +
  plot_layout(widths = c(4, 5), axes = "collect_x", axis_titles = "collect_x")

print(fig3_cell_depth_v2)
save_fig(fig3_cell_depth_v2, "Fig_3A_cell_depth_v2", w = 350/96, h = 200/96, device = "pdf")

# ------------------
# --- Version 3 (IMARIS layers) ---
# ------------------

p_depth_layers <- df_depth %>%
  filter(layer %in% c("L2_3", "L4", "L5_6")) %>%
  ggplot(aes(1, depth, fill = layer, color = layer)) +
  geom_half_violin(side = "r", alpha = 0.3, width = 1.5, trim = TRUE, color = NA) +
  geom_jitter(width = 0.1, size = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.1, alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_y_reverse(limits = c(max_depth, 0)) +
  facet_wrap(~pat, ncol = 3, scales = "fixed", strip.position = "bottom") +
  scale_fill_manual(values = c(L2_3 = "grey80", L4 = "grey50", L5_6 = "grey10")) +
  scale_color_manual(values = c(L2_3 = "grey80", L4 = "grey50", L5_6 = "grey10")) +
  labs(x = "patient", y = "distance from pia (µm)") +
  theme_classic(base_size = 12) +
  theme(
    strip.text = element_text(size = 6),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

p_depth_patients <- df_depth %>%
  filter(layer %in% c("L2_3", "L4", "L5_6")) %>%
  ggplot(aes(1, depth, fill = pat, color = pat)) +
  geom_half_violin(aes(group = layer), side = "r", alpha = 0.3, trim = TRUE, color = NA) +
  geom_jitter(width = 0.1, size = 0.2, alpha = 0.7) +
  geom_boxplot(aes(group = layer), width = 0.5, alpha = 0.3, size = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_y_reverse(limits = c(max_depth, 0), position = "right") +
  facet_wrap(~pat, ncol = 3, scales = "fixed", strip.position = "bottom") +
  scale_fill_manual(values = highlight_colors) +
  scale_color_manual(values = highlight_colors) +
  labs(x = "patient", y = "distance from pia (µm)") +
  theme_classic(base_size = 12) +
  theme(
    strip.text = element_text(size = 6),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

fig3_cell_depth_v3 <- (p3_wfa_dcf | p4_clusters | p_depth_layers) +
  plot_layout(widths = c(4, 6, 7), axes = "collect_x", axis_titles = "collect_x")

print(fig3_cell_depth_v3)
save_fig(fig3_cell_depth_v3, "Fig_3A_cell_depth_v3", w = 550/96, h = 200/96, device = "pdf")
```